Select-based random access \\ to variable-byte encodings

### 2020-04-25 gradu/gradu.tex
Adds slightly more text

\keywords{ulkoasu, tiivistelmä, lähdeluettelo}
\additionalinformation{\translate{\track}}

@@

\classification{\protect{\ \\
\  Information systems $\rightarrow$ Data management systems  $\rightarrow$ Data structures $\rightarrow$ Data layout $\rightarrow$ Data compression \  \\
\  probably $\rightarrow$ something $\rightarrow$ else $\rightarrow$ too \\
}}

>>>>

\keywords{compression, data structure}
\additionalinformation{\translate{\track}}

@@

\classification{\protect{\ \\
\  Information systems $\rightarrow$ Data management systems  $\rightarrow$ Data structures $\rightarrow$ Data layout $\rightarrow$ Data compression \  
}}

### 2020-04-25 gradu/msc_english_contents.tex
Adds slightly more text

A pseudo code for VB encoding with block length 7 is shown in Figure~\ref{vbyte_enc}. \textsc{Prepend} adds an element to the beginning of the list and 
\textsc{extend} adds all the elements of the second list to the end of the first list. The block length can be changed by replacing 128 with $2^b$, where $b$ is the desired block length.

\begin{figure}[ht]
\centering
  \begin{minipage}{0.5\linewidth}
\begin{algorithmic}[H]
\Function{VBEncodeNumber}{$n$}
\State $bytes\gets $ list
\While{true}
\State \Call{prepend}{bytes,$n \bmod 128$}
\If{$n < 128$} \State break \EndIf
\State $n\gets n$ div $128$
\EndWhile
\State $bytes$[\Call{len}{$bytes$}-1] += $128$
\State \textbf{return} bytes
\EndFunction
\medskip
\medskip
\Function{VBEncode}{$numbers$}
\State $bytestream\gets $ list
\ForEach {$n \in numbers$}
\State $bytes \gets$ \Call{VBEncodeNumber}{$n$}
\State \Call{extend}{$bytestream$,$bytes$}
\EndFor
\State \textbf{return} bytestream
\EndFunction

@@

To use rank or select in indexing, the set 1 bits in B should reflect to the element locations in the encoded data. For most compression algorithms, this requires B to be 
created in addition to the existing data and the length of B usually 
has to be close to the length of the data, because encoded element lengths vary. VB encoding has several advantages with search and rank: the data is compressed in 
blocks of equal length which significantly shrinks the length of B. Moreover the bit array C formed from the continuation bits already stores the locations of items and works
as the needed indexing array. In this case, $rank_1$(C, i) would give the sum of end blocks before the i-th index and $select_1$(C, i) would give the location of the ending block 
of i-th compressed element.

@@

The data structure to support $rank_v$ has two layers. The first layer is the superblock array. For every 512th bit, the number of 1's from the beginning of the array is stored to the superblock array.
The second layer is the relative count block. For every 64th bit inside a superblock, it stores the number of 1's since the start of this superblock. To calculate rank(i), the superblock index is 
s = i/512 and relative count block index is r = i-s / 64, both being integer divisions. Then number of 1 bits from index r to i are calculated from the original bit array. All three values added 
together add up to rank(i). $Rank_{v5}$ is a lighter structure: it's superblock size is 2048 and relative counts are taken for every 384th bit. This causes the final bit calculation to be more costly, 
but reduces memory requirement to a quarter.

The $rank$ data structures are based on \citep{Vig08}. A $rank_v$ superblock covers 512 bits and is divided to 512/64 = 8 relative blocks. This means a relative block needs to be able to store a number up to 512 in itself, which takes $log_2$(512) = 9 bits. 
The first relative block always equals to 0, because a relative block counts the number of 1 bits from the start of the superblock. Therefore the remaining 7 relative blocks take 63 bits and fit into one 
64-bit word. The $rank_{v5}$ superblock covers 2048 bits and is divided to 6 relative blocks of 384 bits, requiring $log_2$(2048) = 11 bits each. The first relative block again equals to 0, so the rest 
of the blocks fit to a 64-bit word.

The $select$ data structure works in a similar way to rank. The index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
relative to the superblock. With similar calculations, the location of closest 64th set bit is calculated and then bit array is iterated until the required amount of bits is reached. The $select_{MCL}$
 is a practical variant of a PAT tree \citep{Cla97}.

@@

\section{DAC via rank}
\label{chap:implementations}
Directly addressable VB codes were first introduced by \citep{Bri09}. In their solution, the chunks are stored in separate arrays by their significance. For each integer, the block 
of the least significant chunk is stored in the first array $A_1$, and its continuation bit to the first bit array $B_1$. Then if the number was stored in multiple chunks, the block of the next chunk is
stored to $A_2$ and its continuation bit to $B_2$ and so on. After the data is set, a $rank$ data structure is built for each bit array. The data structure does not require additional space apart
from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure.

@@

To decode an element from index $i$, first block is fetched from $A_1[i]$. Because each encoded element is composed of at least one block, the first block is obtainable via directly indexing $A_1$. 
With small integers, this is all that is required. If the data has a lot of small integers, the $rank$ method thus has a huge advantage. In addition, the block length of this method can change within 
different array levels. This opens a possibility for further compression optimization.

However if the element was stored in multiple blocks, the index of the next block in $A_2$ is obtained from $i \gets rank_0$($A_1$, $i$). $Rank_0$ 
returns the number of zeros in the bit array preceding index $i$. In other words, it returns the number of elements before $i$ that continue to the next array. 
This in turn means that $i$ is the index of desired element's next block in array $A_2$. This process is repeated until the i-th bit of the bit array of the current level (the continuation bit for the current 
block) is set to 1. The resulting integer is constructed from the blocks of the fetched chunks.

A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8. The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and 
contiuation bit arrays $B_k$.

@@

Figure~\ref{select_pseudo} contains an example of VB decoding with DAC with $select$ and block size 8. 
Different block sizes need extra calculation to get the block location from the byte array. In the example, \textsc{CalculateLength} returns the length of the number in blocks and \textsc{ByteMask}(k) returns 
a bit mask for k bytes.

\begin{figure}[ht]
\centering
\begin{algorithmic}
\State $i \gets $ wanted index
\State $A \gets $ block byte array
\State $B \gets $ continuation bit array
\State $begin \gets \Call{Select}{B, index}$
\State $len \gets \Call{CalculateLength}{begin}$
\State $mask \gets \Call{ByteMask}{len}$
\State $word \gets A[begin]$ \Comment{read a word from the byte array}
\State $word \gets word \mathbin{\&} mask$

\end{algorithmic}
\caption{Pseudo code of DAC with select with block length 8} \label{select_pseudo}
\end{figure}

@@

\section{Elias-Fano encoding}
Direct access via select can also be implemented with Elias-Fano encoding, but the data has to be sorted, which is usually a given in inverted index lists of search engines. In Elias-Fano encoding \citep{Eli74, Vig13}, 
the data is at first split into low and 
high bits. Low bits, the k least significant bits, are stored to their own array. Direct access to the lower bit array is trivial. The higher bits are gap-encoded and then encoded in unary. 
For example, [1,3,5,8,11] becomes [1,2,2,3,3] and then is encoded as unary bit array B = \texttt{0100100100010001}. $Select_1$ has an interesting interaction with gap encoded unary arrays. $k = Select_1(B,i)$ 
returns the location of i-th 1 in B. Up to index $k$, the bit array has had $i$ ones and $i-k$ zeros. The $i$ corresponds to the number of compressed integers in $B$ and every zero in the array 
means an increment of 1 in the numbers. In other words, $i-k$ equals to the integer at position $i$ before gap-encoding. Therefore $select_1(i)$ - i returns the original higher bit number. The split 
between bits is made to reduce the gap sizes.
This approach can be further optimized by finding clusters in the integers and partitioning the indexes \citep{Ott14}.

@@

The following experiments were run on a AMD Phenom(tm) II X6 1055T Processor@2.8Ghz with 64kB+64kB L1 cache, 512kB L2 cache, 6144kB L3 cache and 32GB of DDR3 1333MHz.
The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64) with minimal programs running in the background. The codes were compiled with \texttt{g++-7 -std=c++11 -DNDEBUG -O3} 
and with libraries \texttt{-lsdsl -ldivsufsort -ldivsufsort64}.

@@

Each implementation is compressing and dompressing 64-bit unsigned integers. The $rank$ and $select$ implementations are explained in detail in \autoref{chap:implementations} 
and \autoref{chap:select_impl}, respectively.

\pagebreak
Four differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M).

@@

When the dataset contains mostly small integers, the $rank$ based method by Brisaboa et al. is superior to the proposed $select$ based method. The benefit of using a direct array index instead of a $rank$ 
call for the first block gives a huge advantage. The $rank$ query is also slightly faster than a $select$ query, and therefore the dataset needs to have a large portion of longer elements. This can be seen 
from figure~\ref{graph:data_comparison}.

@@

TODO:
 - see how cache matters

 - figure out why data size matters

@@

 - find ACL classifications

 - write keywords

 - remove lists from algorithms and data sets in results, explain whe data sets were roughly chosen like that

 - a real life situation where to use subarray

 - explain rank/select in introduction

 - 1.1 research questions

>>>>

A pseudo code for VB encoding with block length 7 is shown in Figure~\ref{vbyte_enc}. \textsc{Prepend} adds an element to the beginning of the list and 
\textsc{extend} adds all the elements of the second list to the end of the first list. The block length can be changed by replacing 128 with $2^b$, where $b$ is the desired block length. Since the block 
length is 7, the continuation bit is added in row 8 by adding $2^8$ = 128 to the last block.

\begin{figure}[ht]
\centering
  \begin{minipage}{0.5\linewidth}
\begin{algorithmic}[1]
\Function{VBEncodeNumber}{$n$}
\State $bytes\gets $ list
\While{true}
\State \Call{prepend}{bytes,$n \bmod 128$}
\If{$n < 128$} \State break \EndIf
\State $n\gets n$ div $128$
\EndWhile
\State $bytes$[\Call{len}{$bytes$}-1] += $128$
\State \textbf{return} bytes
\EndFunction
\medskip
\medskip
\Function{VBEncode}{$numbers$}
\State $bytestream\gets $ list
\ForEach {$n \in numbers$}
\State $bytes \gets$ \Call{VBEncodeNumber}{$n$}
\State \Call{extend}{$bytestream$,$bytes$}
\EndFor
\State \textbf{return} bytestream
\EndFunction

@@

To use $rank$ or $select$ in indexing, the set 1 bits in the bit array should reflect to the element locations in the encoded data. For most compression algorithms, this requires the bit array to be 
created in addition to the existing data. The length of the bit array in these cases usually has to be close to the length of the data, because encoded element lengths vary. This spatial increment makes 
the compression ineffective and therefore $rank$ and $select$ cannot be applied to any given compression algorithm.

VB encoding has several advantages with $search$ and $rank$: the data is compressed in 
blocks of equal length which significantly shrinks the length of B. Moreover the bit array C formed from the continuation bits already stores the locations of items and works
as the needed indexing array. In this case, $rank_1$(C, i) would give the sum of end blocks before the i-th index and $select_1$(C, i) would give the location of the ending block 
of i-th compressed element.

@@

The data structure to support $rank_v$ has two layers. The first layer is the superblock array. For every 512th bit, the number of 1's from the beginning of the array is stored to the superblock array.
The second layer is the relative count block. For every 64th bit inside a superblock, it stores the number of 1's since the start of this superblock.

For $rank$($i$), the superblock index $s$ = $i$ / 512 and relative count block index $r$ = ($i$ - $s$) / 64 need to be calculated, both division operations being integer divisions. Then number of 1 bits 
from index $r$ to $i$ are calculated from the original bit array. All three values added 
together add up to $rank$($i$). $Rank_{v5}$ is a lighter structure: it's superblock size is 2048 and relative counts are taken for every 384th bit. This causes the final bit calculation to be more costly, 
but reduces memory requirement to a quarter.

The $rank$ data structures are based on \citep{Vig08}. A $rank_v$ superblock covers 512 bits and is divided to 512/64 = 8 relative blocks. This means a relative block needs to be able to store a number up to 512 in itself, which takes $log_2$(512) = 9 bits. 
The first relative block always equals to 0, because a relative block counts the number of 1 bits from the start of the superblock. Therefore the remaining 7 relative blocks take 63 bits and fit into one 
64-bit word. For every 512 bits, this implementation requires 64 bits for the superblock and another 64 blocks for the relative blocks, resulting in additional 128/512 = 25\% space.

The $rank_{v5}$ superblock covers 2048 bits and is divided to 6 relative blocks of 384 bits, requiring $log_2$(2048) = 11 bits each. The first relative block again equals to 0, so the rest 
of the blocks fit to a 64-bit word. The $rank_{v5}$ implementation requires an additional (64+64)/2048 = 6.25\% space.

The $select$ data structure works in a similar way to rank. The index location of every 4096th set bit is stored in the superblock and location of every 64th set bit is stored 
relative to the superblock. With similar calculations, the location of closest 64th set bit is calculated and then bit array is iterated until the required amount of bits is reached. The $select_{MCL}$
 is a practical variant of a PAT tree \citep{Cla97}. The memory requirement of the $select$ implementation depends on the portion of 1 bits in the bit array.

@@

\section{DAC via rank}
\label{chap:implementations}
Directly addressable VB codes were first introduced by \citep{Bri09}. In their solution, the chunks are stored in separate arrays by their significance. For each integer, the block 
of the least significant chunk is stored in the first array $A_1$, and its continuation bit to the first bit array $B_1$. Then if the number was stored in multiple chunks, the block of the next chunk is
stored to $A_2$ and its continuation bit to $B_2$ and so on. After the data is set, a $rank$ data structure is built for each bit array. The data structure does not require additional space apart
from the $rank$ structure. Figure~\ref{bris_ds} contains a visualization of the data structure. The C array is the original VB encoded data, which is split into separate block arrays $A_k$ and 
contiuation bit arrays $B_k$.

@@

To decode an element from index $i$, first block is fetched from $A_1[i]$. Because each encoded element is composed of at least one block, the first block is obtainable via directly indexing $A_1$. 
With small integers, this is all that is required. If the data has a lot of small integers, the $rank$ method thus has a huge advantage. In addition, the block length of this method can change within 
different array levels which opens a possibility for further compression optimization.

However if the element was stored in multiple blocks, the index of the next block in $A_2$ is obtained from $i \gets rank_0$($A_1$, $i$). $Rank_0$ 
returns the number of zeros in the bit array preceding index $i$. In other words, it returns the number of elements before $i$ that continue to the next array. 
This in turn means that $i$ is the index of desired element's next block in array $A_2$. This process is repeated until the i-th bit of the bit array of the current level (the continuation bit for the current 
block) is set to 1. The resulting integer is constructed from the blocks of the fetched chunks. A pseudo code example of DAC with $rank$ is shown in Figure~\ref{bris_pseudo} with block length of 8.

@@

Figure~\ref{select_pseudo} contains an example of VB decoding with DAC with $select$ and block size 8. 
Different block sizes need extra calculation to get the block location from the byte array. In the example, \textsc{CalculateLength} returns the length of the number in blocks and \textsc{ByteMask}(k) returns 
a bit mask for k bytes. In line 8, a long word is read from the memory starting at memory address $A$[$begin$].

\begin{figure}[ht]
\centering
\begin{minipage}{0.5\linewidth}
\begin{algorithmic}[1]
\State $i \gets $ wanted index
\State $A \gets $ block byte array
\State $B \gets $ continuation bit array
\State $begin \gets \Call{Select}{B, index}$
\State $len \gets \Call{CalculateLength}{begin}$
\State $mask \gets \Call{ByteMask}{len}$
\State $word \gets A[begin]$
\State $word \gets word \mathbin{\&} mask$

\end{algorithmic}
\end{minipage}
\caption{Pseudo code of DAC with select with block length 8} \label{select_pseudo}
\end{figure}

@@

\section{Elias-Fano encoding}
Direct access via select can also be implemented with Elias-Fano encoding, but the data has to be sorted, which is usually a given in inverted index lists of search engines. In Elias-Fano encoding \citep{Eli74, Vig13}, 
the data is at first split into low and 
high bits. Low bits, the k least significant bits, are stored to their own array. Direct access to the lower bit array is trivial. The higher bits are gap-encoded and then encoded in unary. 
For example, [1,3,5,8,11] becomes [1,2,2,3,3] and then is encoded as unary bit array B = \texttt{0100100100010001}.

$Select_1$ has an interesting interaction with gap encoded unary arrays. $k = Select_1(B,i)$ 
returns the location of i-th 1 in B. Up to index $k$, the bit array has had $i$ ones and $i-k$ zeros. The $i$ corresponds to the number of compressed integers in $B$ and every zero in the array 
means an increment of 1 in the numbers. In other words, $i-k$ equals to the integer at position $i$ before gap-encoding. Therefore $select_1(i)$ - i returns the original higher bit number. The split 
between bits is made to reduce the gap sizes.
This approach can be further optimized by finding clusters in the integers and partitioning the indexes \citep{Ott14}.

@@

The following experiments were run on a AMD Phenom(tm) II X6 1055T Processor@2.8Ghz with 64kB+64kB L1 cache, 512kB L2 cache, 6144kB L3 cache and 32GB of DDR3 1333MHz.
The computer runs Ubuntu 14.04.4 (3.13.0-91-generic x86\_64) with minimal programs running in the background. The implementations were written in C++ and compiled with 
\texttt{g++-7 -std=c++11 -DNDEBUG -O3} 
and with libraries \texttt{-lsdsl -ldivsufsort -ldivsufsort64}.

@@

Each implementation is compressing and decompressing 64-bit unsigned integers. The $rank$ and $select$ implementations are explained in detail in \autoref{chap:implementations} 
and \autoref{chap:select_impl}, respectively.

Four differently constructed datasets were used and three different sizes of dataset were used (5M, 50M, 500M). The content of the datasets were chosen as such to see how different implementations 
work with different sized data.

@@

When the dataset contains mostly small integers, the $rank$ based method by Brisaboa et al. is superior to the proposed $select$ based method. The benefit of using a direct array index instead of a $rank$ 
call for the first block gives a huge advantage. The $rank$ query is also slightly faster than a $select$ query, and therefore the dataset needs to have a large portion of longer elements until $select$ 
performs better. This can be seen from figure~\ref{graph:data_comparison}.

@@

TODO:

@@

 - remove lists from algorithms and data sets in results?

 - a real life situation of where to use subarray

 - list research questions somewhere? (introduction 1.1?)
